
# Exploratory Data Analysis with Unsupervised Machine Learning

## Very helpful youtube explanations
- I think I've mentioned [StatQuest](https://www.youtube.com/@statquest) at almost every bookclub meeting, but these are the clearest and most intuitive explanations I've found regarding biostatistics. 

[PCA concepts](https://www.youtube.com/watch?v=HMOI_lkzW08)

[PCA step-by-step](https://www.youtube.com/watch?v=FgakZw6K1QQ)

[Even more PCA tips](https://www.youtube.com/watch?v=oRvgq966yZg)

[MDS](https://www.youtube.com/watch?v=GEn-_dAyYME)

[tSNE](https://www.youtube.com/watch?v=NEaUSP4YerM)

[Not a StatQuest video, but useful in explaining NNMF in the context of cancer mutation signatures](https://www.youtube.com/watch?v=1_43nCptm44) Recommend skipping to 16:45.

## Chapter 4 Exercises

Data set and packages

```{r, warning = FALSE}
expFile=system.file("extdata", "leukemiaExpressionSubset.rds", package="compGenomRData") 

mat=readRDS(expFile) 

library(pheatmap) 
library(cluster) 
library(fastICA)
library(Rtsne)
```

```{r, collapse = TRUE}
## 1
x <- scale(mat)
y <- scale(log2(mat))

boxplot(mat) 
boxplot(x) 
boxplot(y)

```


```{r, collapse = TRUE}
## 2 
annotation_col = data.frame(LeukemiaType =substr(colnames(mat),1,3))
rownames(annotation_col)=colnames(mat)

pheatmap(mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale = "none",clustering_method="ward.D2",
         clustering_distance_cols="euclidean")

pheatmap(mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale = "none",clustering_method="ward.D",
         clustering_distance_cols="euclidean")

### IMO this one looks the bet to me
pheatmap(x,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale = "none",clustering_method="ward.D",
         clustering_distance_cols="euclidean")

pheatmap(y,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale = "none",clustering_method="ward.D",
         clustering_distance_cols="euclidean")

pheatmap(mat,show_rownames=FALSE,show_colnames=FALSE,
         annotation_col=annotation_col,
         scale = "column",clustering_method="ward.D",
         clustering_distance_cols="euclidean")

```


```{r, collapse = TRUE}
## 3  number of clusters 
set.seed(101) 
pamclu=cluster::pam(t(mat),k=5)
plot(silhouette(pamclu),main=NULL)


# even when mat is changed with scaled data it looks the same

Ks=sapply(2:7,
          function(i) 
            summary(silhouette(pam(t(mat),k=i)))$avg.width) 
plot(2:7,Ks,xlab="k",ylab="av. silhouette",type="b", pch=19)
```



```{r, collapse = TRUE}
## 4 
 set.seed(101) 
 
 # define the clustering function 
 pam1 <- function(x,k) list(cluster = pam(x,k, cluster.only=TRUE)) 
 
 #calculate the gap statistic
 pam.gap= clusGap(t(mat), FUN = pam1, K.max = 8,B=50)
 
 #plot the gap statistic accross k values
 plot(pam.gap, main = "Gap statistic for the 'Leukemia' data")

```


```{r, collapse = TRUE}
## 5 

#data all together doesn't tell you anything useful 

plot(mat, pch = 19, col=as.factor(annotation_col$LeukemiaType)) 
princomp(x)
screeplot(princomp(x))
```


```{r, collapse = TRUE}
#create the subset of the data with two genes only. notice that we transpose the matrix so samples are on the columns

par(mfrow=c(1,2))
sub.mat=t(mat[rownames(mat) %in% c("ENSG00000100504","ENSG00000105383"),])
```

```{r, collapse= TRUE}
#create the subset of the data with two genes only. notice that we transpose the matrix so samples are on the columns

plot(scale(mat[rownames(mat)=="ENSG00000100504",]),
     scale(mat[rownames(mat)=="ENSG00000105383",]),
     pch=19,
     ylab="CD33 (ENSG00000105383)",
     xlab="PYGL (ENSG00000100504)",
     col=as.factor(annotation_col$LeukemiaType),
     xlim=c(-2,2),ylim=c(-2,2))


#create the legend for the Leukemia types
legend("bottomright",
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)
```


```{r, collapse= TRUE}
# calculate the PCA only for our genes and all the samples
pr=princomp(scale(sub.mat))

#screeplot of PCA
screeplot(pr)

#plot the direction of eigenvectors
#pr$loadings returned by princomp has the eigenvectors

arrows(x0=0, y0=0, x1 = pr$loadings[1,1], 
       y1 = pr$loadings[2,1],col="pink",lwd=3)
arrows(x0=0, y0=0, x1 = pr$loadings[1,2], 
       y1 = pr$loadings[2,2],col="gray",lwd=3)

```

```{r, collapse = TRUE}
#plot the samples in the new coordinate system
plot(-pr$scores,pch=19,
     col=as.factor(annotation_col$LeukemiaType),
     ylim=c(-2,2),xlim=c(-4,4))

#plot the new coordinate basis vectors

arrows(x0=0, y0=0, x1 =-2, 
       y1 = 0,col="pink",lwd=3)
arrows(x0=0, y0=0, x1 = 0, 
       y1 = -1,col="gray",lwd=3)
```



```{r, collapse = TRUE}
## 6 
par(mfrow=c(1,2))
d=svd(scale(mat)) # apply SVD
assays=t(d$u) %*% scale(mat) # projection on eigenassays
plot(assays[1,],assays[2,],pch=19,
     col=as.factor(annotation_col$LeukemiaType))
pr=prcomp(t(mat),center=TRUE,scale=TRUE) # apply PCA on transposed matrix

# plot new coordinates from PCA, projections on eigenvectorssince the matrix is transposed eigenvectors represent 

plot(pr$x[,1],pr$x[,2],col=as.factor(annotation_col$LeukemiaType))
```



```{r, collapse = TRUE}
## 7
# apply ICA
ica.res=fastICA(t(mat),n.comp=2) 

# plot reduced dimensions
plot(ica.res$S[,1],ica.res$S[,2],col=as.factor(annotation_col$LeukemiaType)) 

```



```{r, collapse = TRUE}
## 8 
set.seed(42) # Set a seed if you want reproducible results
tsne_out <- Rtsne(t(mat),perplexity = 10) # Run TSNE
                                          # image(t(as.matrix(dist(tsne_out$Y))))  
                                          # Show the objects in the 2D tsne representation
                                          
plot(tsne_out$Y,col=as.factor(annotation_col$LeukemiaType),pch=19)
legend("bottomleft", # create the legend for the Leukemia types
       legend=unique(annotation_col$LeukemiaType),
       fill =palette("default"),
       border=NA,box.col=NA)
```


## Meeting Videos

### Cohort 1

`r knitr::include_url("https://www.youtube.com/embed/URL")`

<details>

<summary>Meeting chat log</summary>

    LOG

</details>
